---
title: "Analysis of US Smoking Data (2011-2017)"
author: "Peter J. Ehmann"
date: 'Due: 5/6/2019'
output: html_document
---

```{r setup, include = FALSE}
library(broom)
library(jsonlite)
library(knitr)
library(rvest)
library(tidyverse)
library(xml2)
opts_chunk$set(echo = FALSE)
```

<br>

# Part 1 - Acquire, clean, and format data (export as .csv).

<br>

#### US state smoking data from the Centers for Disease Control (CDC) is availalbe through the Socrata API. The JSON-formatted data for smoking rates by state from 2011-2017 can be accessed from https://chronicdata.cdc.gov/Survey-Data/Table-for-STATE-System-Current-Cigarette-Use-Among/xmxq-jxrr. There are a total of 371 JSON objects (rows) in the original dataset (53 states [includes DC, PR, GU] x 7 rows per state). A few columns of interest (renamed) are shown below. The smoking rate (%) and year are nested as 7x2 tibbles for each state.

```{r smoking_rates, warning = FALSE}
(
smoking_rates <- "https://chronicdata.cdc.gov/resource/gx47-p4ij.json" %>% 
  fromJSON() %>% 
  as_tibble() %>% 
  rename(state = locationdesc, abbr = locationabbr, year = year, percent = data_value) %>% 
  select(state, abbr, year, percent) %>% 
  group_by(state, abbr) %>% 
  nest(.key = smoking_rate) %>% 
  filter(abbr != "PR", 
         abbr != "GU")
  )
```

<br>

#### US state tobacco taxation (Cigarette - $ per pack) data from 1995-2019 is also available from the CDC through the Socrata API. Information about the dataset can be obtained at https://chronicdata.cdc.gov/Legislation/CDC-STATE-System-Tobacco-Legislation-Tax/2dwv-vfam. The data can be accessed as JSON and filtered before retrieving it since there are a total of 152,192 rows in the complete dataset. I extracted taxation levels in 2011 and 2017, and calculated a change score for each state.

```{r tax}
(
  tax <- "https://chronicdata.cdc.gov/resource/7uzt-fasa.json?Year=2011%20AND%20MeasureDesc=%27Cigarette%27%20OR%20Year=2017%20AND%20MeasureDesc=%27Cigarette%27" %>% 
    fromJSON() %>% 
    as_tibble() %>% 
    rename(state = locationdesc, tax_value = provisionvalue) %>% 
    filter(state %in% as_vector(smoking_rates[,1])) %>% 
    mutate(tax_value = as.numeric(tax_value)) %>% 
    select(state, year, tax_value) %>% 
    distinct() %>% 
    group_by(state, year) %>% 
    summarize(value = mean(tax_value)) %>% 
    spread(., year, value) %>% 
    rename(tax_2011 = `2011`, tax_2017 = `2017`) %>% 
    mutate(delta_tax = tax_2017 - tax_2011)
)
```

<br>

#### US state preemption data from 1995-2019 is also available from the CDC through the Socrata API. Information about the dataset can be obtained at https://chronicdata.cdc.gov/Legislation/CDC-STATE-System-Tobacco-Legislation-Preemption/xsta-sbh5. The data can be accessed as JSON and filtered before retrieving it since there are a total of 107,656 rows in the complete dataset. There are 4 main types of preemptions: (1) advertising, (2) licensure, (3) smokefree indoor air, and (4) youth access. I extracted a count for each state {Min = 0, Max = 4}.

```{r preemptions}
(
  preemptions <- "https://chronicdata.cdc.gov/resource/ksh7-354r.json?year=%272017%27%20AND%20provisionaltvalue=%271%27%20OR%20year=%272017%27%20AND%20provisionaltvalue=%272%27%20LIMIT%2050000" %>% 
    fromJSON() %>% 
    as_tibble() %>% 
    rename(state = locationdesc, preemption_type = measuredesc) %>% 
    filter(state %in% as_vector(smoking_rates[,1])) %>% 
    mutate(value = as.numeric(provisionaltvalue)) %>% 
    group_by(state, preemption_type) %>% 
    summarize(status = mean(value)) %>% 
    mutate(status = ceiling(status) - 1) %>% 
    select(state, status) %>% 
    group_by(state) %>% 
    summarize(num_of_preemptions = sum(status))
)
```

<br>

#### US state smoking restrictions law (smokefree indoor air) data from 1995-2019 is also available from the CDC through the Socrata API. Information about the dataset can be obtained at https://chronicdata.cdc.gov/Legislation/CDC-STATE-System-Tobacco-Legislation-Smokefree-Ind/32fd-hyzc. The data can be accessed as JSON and filtered before retrieving it since there are a total of 575,360 rows in the complete dataset. I extracted restrictions in Bars, Restaurants, and Workplaces in 2017 (Q4). Missing values were assumed to be indicative of 'No restriction'.

```{r indoor_air}
(
  indoor_air <- "https://chronicdata.cdc.gov/resource/ag3f-urcg.json?year=%272017%27%20AND%20measuredesc=%27Private%20Worksites%27%20AND%20provisionaltvalue!=%270%27%20AND%20quarter=%274%27%20AND%20provisionid=%2720%27%20OR%20year=%272017%27%20AND%20measuredesc=%27Bars%27%20AND%20provisionaltvalue!=%270%27%20AND%20quarter=%274%27%20AND%20provisionid=%27234%27%20OR%20year=%272017%27%20AND%20measuredesc=%27Restaurants%27%20AND%20provisionaltvalue!=%270%27%20AND%20quarter=%274%27%20AND%20provisionid=%2737%27" %>% 
    fromJSON() %>% 
    as_tibble %>% 
    rename(state = locationdesc, place = measuredesc, restriction = provisionvalue) %>% 
    filter(state %in% as_vector(smoking_rates[,1])) %>% 
    select(state, place, restriction) %>% 
    spread(., place, restriction) %>% 
    rename(bar_restrictions = `Bars`, work_restrictions = `Private Worksites`, restaurant_restrictions = `Restaurants`) %>% 
    replace_na(list(bar_restrictions = "None", work_restrictions = "None", restaurant_restrictions = "None"))
)
```

<br>

#### US funding (per capita) data from 1991-2014 is also available from the CDC through the Socrata API. Information about the dataset can be obtained at https://chronicdata.cdc.gov/Funding/University-of-Illinois-at-Chicago-Health-Policy-Ce/vw7y-v3uk. The data can be accessed as JSON and filtered before retrieving it since there are a total of 12,514 rows in the complete dataset. I extracted FUNDING & EXPENDITURES using 'Total Per Capita' in 2014 (most recent data), and calculated the percent of total funding utilized that year.

```{r funding}
(
  funding <- "https://chronicdata.cdc.gov/resource/cfhu-b5vt.json?source=%27Total%20Per%20Capita%27%20AND%20year=%272014%27" %>% 
    fromJSON() %>% 
    as_tibble() %>% 
    rename(state = locationdesc, funding = data_value, funding_type = measuredesc) %>% 
    filter(state %in% as_vector(smoking_rates[,1])) %>% 
    select(state, funding, funding_type) %>% 
    group_by(state, funding_type) %>% 
    spread(., funding_type, funding) %>% 
    mutate(expenditures_per_capita = as.numeric(`Expenditures`), 
           funding_per_capita = as.numeric(`Appropriations/Grants`)) %>% 
    select(state, funding_per_capita, expenditures_per_capita) %>% 
    mutate(percent_funding_used = expenditures_per_capita*100/funding_per_capita)
)
```

<br>

#### US state Medicaid data from 2008-2019 is also available from the CDC through the Socrata API. Information about the dataset can be obtained at https://chronicdata.cdc.gov/Cessation-Coverage-/Medicaid-Coverage-Of-Cessation-Treatments-And-Barr/ntaa-dtex/data. The data can be accessed as JSON and filtered before retrieving it since there are a total of 27,132 rows in the complete dataset. For the last quarter of 2017, I extracted data regarding 'fee for service plans' and 'managed care plans'.

```{r medicaid}
(
  medicaid <- "https://chronicdata.cdc.gov/resource/ufr7-5jfh.json?year=%272017%27%20AND%20quarter=%274%27%20AND%20measure=%27Medicaid%20Coverage%20of%20Cessation%20Treatments%27" %>% 
    fromJSON() %>% 
    as_tibble() %>% 
    rename(state = locationdesc) %>% 
    filter(state %in% as_vector(smoking_rates[,1])) %>% 
    filter(fee_for_service_plans_altvalue != 3, 
           fee_for_service_plans_altvalue != 4, 
           managed_care_plans_altvalue != 3, 
           managed_care_plans_altvalue != 4, 
           submeasure == "Comprehensive Medicaid Coverage of Treatments") %>% 
    select(state, fee_for_service_plans, managed_care_plans) %>% 
    distinct(state, fee_for_service_plans, managed_care_plans) %>% 
    arrange(state)
)
```

<br>

#### US state quiline usage (incoming calls per 10,000 state population) data from 2010-2016 is also available from the CDC through the Socrata API. Information about the dataset can be obtained at https://chronicdata.cdc.gov/Quitline/Quitline-Service-Utilization-2010-To-Present/equ4-92qe. The data can be accessed as JSON and filtered before retrieving it since there are a total of 87,240 rows in the complete dataset. I extracted 'number of incoming calls per 10,000 state population' in 2016 (most recent data), and calculated the average over all 4 quarters of that year.

```{r quitline}
(
  quitline <- "https://chronicdata.cdc.gov/resource/ew86-dx3i.json?variable=%27Incoming%20Calls%20per%2010,000%20State%20Population%27%20AND%20year=%272016%27" %>% 
    fromJSON() %>% 
    as_tibble() %>% 
    rename(state = locationdesc, quitline_percent = value) %>% 
    filter(state %in% as_vector(smoking_rates[,1])) %>% 
    mutate(quitline_percent = as.numeric(quitline_percent)) %>% 
    select(state, quitline_percent) %>% 
    group_by(state) %>% 
    summarise(quitline_calls_per_10000 = mean(quitline_percent))
)
```

<br>

#### Information about US state region classifications can be scraped from the main list in the HTML of https://simple.wikipedia.org/wiki/List_of_regions_of_the_United_States. The div, ul, and il tags can be used to extract each state's region and division (sub-region).

```{r scrape_wiki_html}
(
states_html <- read_html("https://simple.wikipedia.org/wiki/List_of_regions_of_the_United_States")
)

states <- states_html %>% 
  html_nodes(xpath = "//div/ul/li/ul/li/ul/li") %>% 
  html_text()

divisions <- states_html %>% 
  html_nodes(xpath = "//div/div/div/div/ul/li/ul/li") %>% 
  html_text()

regions <- states_html %>% 
  html_nodes(xpath = "//div/div/div/div/ul/li") %>% 
  html_text()
```

```{r extract_division}
division_names <- divisions %>% 
  str_extract(., "\\((.*)\\)") %>% 
  str_extract(., "[\\w+\\s?]+")

division_counts <- divisions %>% 
  str_count(., "\n")

division_list = ""
for (i in 1:length(division_names)) {
  division_list <- c(division_list, (rep(division_names[i], division_counts[i])))
}
```

```{r extract_region}
region_names <- regions %>% 
  .[1:4] %>% 
  str_extract(., "\\((.*)\\)") %>% 
  str_extract(., "[\\w+\\s?]+")

region_counts <- regions %>% 
  .[1:4] %>% 
  str_count(., "\n")

divisions_per_region <- regions %>% 
  .[1:4] %>% 
  str_count(., "Division")

region_list = ""
for (i in 1:length(region_names)) {
  region_list <- c(region_list, (rep(region_names[i], region_counts[i]-divisions_per_region[i])))
}
```

```{r state_table}
state_table <- as_tibble(data.frame(matrix(nrow = length(states), ncol = 3)))
colnames(state_table) <- c("state", "region", "division")
state_table[,1] <- states
state_table[,2] <- region_list[-1]
state_table[,3] <- division_list[-1]
state_table %>% arrange(state)
```

<br>

#### The datasets are merged by STATE and the nested tibbles are unnested.

```{r merge_data}
df_clean <- merge(smoking_rates, tax, by = "state") %>% 
  merge(., preemptions, by = "state") %>% 
  merge(., indoor_air, by = "state", all = TRUE) %>% 
  merge(., funding, by = "state") %>% 
  merge(., medicaid, by = "state") %>% 
  merge(., quitline, by = "state") %>% 
  merge(., state_table, by = "state") %>% 
  replace_na(list(bar_restrictions = "None", work_restrictions = "None", restaurant_restrictions = "None")) %>% 
  unnest()
```

<br>

#### I also extracted simple linear regression parameters (b0 = intercept @ t=0 [2011], b1 = slope) and residuals (observed - expected) at three different levels (state, division, region) and added them to the dataset.

```{r fxn_extract_regression_int_slope}
regression <- function(df) {
  lm(percent ~ year, data = df)
}
parameter <- function(df, choice) {
  tidy(df)[[choice,"estimate"]]
}
```

```{r state_regression}
df_clean <- df_clean %>% 
  mutate(year = as.numeric(year) - 2011) %>% 
  group_by(state) %>% 
  nest() %>% 
  mutate(lm_fit = map(data, regression)) %>% 
  mutate(state_int = as.numeric(map2(lm_fit, 1, parameter))) %>% 
  mutate(state_slope = as.numeric(map2(lm_fit, 2, parameter))) %>% 
  select(-lm_fit) %>% 
  unnest() %>% 
  mutate(percent = as.numeric(percent)) %>% 
  mutate(state_resid = percent - (state_int + state_slope*year)) %>% 
  mutate(year = year + 2011)
```

```{r division_regression}
df_clean <- df_clean %>% 
  mutate(year = year - 2011) %>% 
  group_by(division) %>% 
  nest() %>% 
  mutate(lm_fit = map(data, regression)) %>% 
  mutate(division_int = as.numeric(map2(lm_fit, 1, parameter))) %>% 
  mutate(division_slope = as.numeric(map2(lm_fit, 2, parameter))) %>% 
  select(-lm_fit) %>% 
  unnest() %>% 
  mutate(division_resid = percent - (division_int + division_slope*year)) %>% 
  mutate(year = year + 2011)
```

```{r region_regression}
df_clean <- df_clean %>% 
  mutate(year = year - 2011) %>% 
  group_by(region) %>% 
  nest() %>% 
  mutate(lm_fit = map(data, regression)) %>% 
  mutate(region_int = as.numeric(map2(lm_fit, 1, parameter))) %>% 
  mutate(region_slope = as.numeric(map2(lm_fit, 2, parameter))) %>% 
  select(-lm_fit) %>% 
  unnest() %>% 
  mutate(region_resid = percent - (region_int + region_slope*year)) %>% 
  mutate(year = year + 2011)
```

<br>

#### The last step is to export the data as a .csv file which is posted in the GitHub repository.

```{r write_csv}
# uncomment line 306 to clear the global enviroment of all variables
df_clean %>% select(state, abbr, region, division, year, percent, state_int, state_slope, state_resid, division_int, division_slope, 
                    division_resid, region_int, region_slope, region_resid, tax_2011, tax_2017, delta_tax, num_of_preemptions, bar_restrictions, 
                    work_restrictions, restaurant_restrictions, funding_per_capita, expenditures_per_capita, percent_funding_used, 
                    fee_for_service_plans, managed_care_plans, quitline_calls_per_10000) %>% 
  write_csv("smoking_data.csv")
# rm(list = ls(all = TRUE))
```

<br>

# Part 2 - Data visualization and statistics.

<br>

#### The cleaned data can be retreived from the GitHub repository at https://raw.githubusercontent.com/peter-ehmann/Data_Wrangling_Project_S2019/master/smoking_data.csv.

```{r import_clean_data, message = FALSE}
(
  data <- read_csv("https://raw.githubusercontent.com/peter-ehmann/Data_Wrangling_Project_S2019/master/smoking_data.csv")
)
```

<br>

### First, we will look at differences in smoking laws and policies by region.

<br>

#### TAXATION. From the bar plot, it is apparent that Northeast states tax cigarettes more and had the greatest increase from 2011-2017. The South had the lowest tax rates and also the smallest increase in tax from 2011-2017.

```{r tax_analysis}
data %>% distinct(state, region, tax_2011, tax_2017, delta_tax) %>% 
  group_by(region) %>% 
  summarize(tax_2011_avg = mean(tax_2011), 
            tax_2017_avg = mean(tax_2017), 
            delta_tax_avg = mean(delta_tax)) %>% 
  gather(., variable, value, -region) %>% 
  ggplot(., mapping = aes(x = variable, y = value, fill = region)) + 
    geom_col(position = position_dodge()) + 
    xlab("") + 
    ylab("Tax per carton of Cigarettes ($)") + 
    ggtitle("Taxation of Cigarettes by US Region") + 
    scale_x_discrete(labels = c("delta_tax_avg" = "Change", "tax_2011_avg" = "2011", "tax_2017_avg" = "2017"))
```

<br>

#### PREEMPTIONS. The bar graph shows wide variability in the number of states with preemptions within each region.

```{r preemption_analysis}
data %>% distinct(state, region, num_of_preemptions) %>% 
  group_by(region, num_of_preemptions) %>% 
  tally() %>% 
  spread(., num_of_preemptions, n) %>% 
  replace_na(list(`0` = 0, `1` = 0, `2` = 0, `3` = 0, `4` = 0)) %>% 
  gather(., num_pre, num_states, -region) %>% 
  ggplot(., mapping = aes(x = num_pre, y = num_states, fill = region)) + 
    geom_col() + 
    xlab("Number of Preemptions") + 
    ylab("Number of States") + 
    ggtitle("Number of Preemptions by Region (Max = 4)")
```

<br>

#### INDOOR AIR. The South appaers to have the most instances of 'None' across bars, restaurants, and workplaces.

```{r indoor_analysis}
indoor_air_analysis <- data %>% 
  distinct(state, region, bar_restrictions, work_restrictions, restaurant_restrictions) %>% 
  group_by(region, bar_restrictions, work_restrictions, restaurant_restrictions) %>% 
  tally() %>% 
  spread(., bar_restrictions, n) %>% 
  replace_na(list(`Banned` = 0, `Designated Areas` = 0, `None` = 0, `Separate Ventilated Areas` = 0)) %>% 
  gather(., bar_restrictions, n, `Banned`:`Separate Ventilated Areas`) %>% 
  spread(., work_restrictions, n) %>% 
  replace_na(list(`Banned` = 0, `Designated Areas` = 0, `None` = 0, `Separate Ventilated Areas` = 0)) %>% 
  gather(., work_restrictions, n, `Banned`:`Separate Ventilated Areas`) %>% 
  spread(., restaurant_restrictions, n) %>% 
  replace_na(list(`Banned` = 0, `Designated Areas` = 0, `None` = 0, `Separate Ventilated Areas` = 0)) %>% 
  gather(., restaurant_restrictions, n, `Banned`:`Separate Ventilated Areas`)

ggplot(indoor_air_analysis, mapping = aes(x = bar_restrictions, y = n, fill = region)) + 
  geom_col() + 
  xlab("") + 
  ylab("Number of States") + 
  ggtitle("Bar Restrictions")

ggplot(indoor_air_analysis, mapping = aes(x = work_restrictions, y = n, fill = region)) + 
  geom_col() + 
  xlab("") + 
  ylab("Number of States") + 
  ggtitle("Work Restrictions")

ggplot(indoor_air_analysis, mapping = aes(x = restaurant_restrictions, y = n, fill = region)) + 
  geom_col() + 
  xlab("") + 
  ylab("Number of States") + 
  ggtitle("Restaurant Restrictions")
```

<br>

#### FUNDING. At least half of Northeast and Midwest states use all the money provided through various funding sources.

```{r funding_analysis}
data %>% distinct(state, region, percent_funding_used) %>% 
  group_by(region) %>% 
  add_tally(percent_funding_used > 99.99) %>% 
  add_tally(n()) %>% 
  mutate(percent = n*100/nn) %>% 
  distinct(region, percent) %>% 
  ggplot(., mapping = aes(x = "", y = percent, fill = region)) + 
    geom_col(position = position_dodge()) + 
    xlab("") + 
    ylab("Percent of States in Region") + 
    ggtitle("States Expending >99.99 % of Funding Provided")
```

<br>

#### MEDICAID. Most states do not provide coverage, and trends are even across all regions.

```{r medicaid_analysis}
medicaid_analysis <- data %>% 
  distinct(state, region, fee_for_service_plans, managed_care_plans) %>% 
  group_by(region, fee_for_service_plans, managed_care_plans) %>% 
  tally() %>% 
  spread(., fee_for_service_plans, n) %>% 
  replace_na(list(`No` = 0, `Not Applicable` = 0, `Yes` = 0)) %>% 
  gather(., fee_for_service_plans, n, `No`:`Yes`) %>% 
  spread(., managed_care_plans, n) %>% 
  replace_na(list(`No` = 0, `Not Applicable` = 0, `Yes` = 0)) %>% 
  gather(., managed_care_plans, n, `No`:`Yes`)

ggplot(medicaid_analysis, mapping = aes(x = fee_for_service_plans, y = n, fill = region)) + 
  geom_col() + 
  xlab("") + 
  ylab("Number of States") + 
  ggtitle("Fee for Service Plans")

ggplot(medicaid_analysis, mapping = aes(x = managed_care_plans, y = n, fill = region)) + 
  geom_col() + 
  xlab("") + 
  ylab("Number of States") + 
  ggtitle("Managed Care Plans")
```

<br>

#### QUITLINE. The higher amount of quitline users may be highly correlated with number of smokers. This is confirmed by the linear regression analysis below.

```{r quitline_analysis}
data %>% distinct(state, region, quitline_calls_per_10000) %>% 
  group_by(region) %>% 
  summarize(mean = mean(quitline_calls_per_10000)) %>% 
  ggplot(., mapping = aes(x = "", y = mean, fill = region)) + 
    geom_bar(width = 1, stat = "identity") + 
    coord_polar("y", start = 0) + 
    theme(axis.text.x = element_blank()) + 
    xlab("") + 
    ylab("") + 
    ggtitle("Percent of Quitline Calls by Region")

summary(
  lm(quitline_calls_per_10000 ~ state_int, data = data %>% distinct(state, state_int, quitline_calls_per_10000))
)
```

<br>

### Now, we will plot out smoking rates from 2011-2017.

<br>

#### A simple line plot grouped by REGION shows a similar linear decrease in '% smokers' for all 4 US regions. However, the MIDWEST/SOUTH have much higher smoking rate for all years than the NORTHEAST/WEST.

```{r region_line_plot}
data %>% select(state, region, year, percent) %>% 
  group_by(region, year) %>% 
  summarize(average = mean(percent)) %>% 
  ggplot(., mapping = aes(x = year, y = average, color = region, group = region)) + 
    geom_point() + 
    geom_line() + 
    xlab("Year") + 
    ylab("Current Cigarette Smokers (%)") + 
    ggtitle("Smoking Rates (%) by US Region") + 
    scale_x_continuous(breaks = pretty(c("2011":"2017"), n = 7))
```

<br>

#### The table below shows approximate smoking rates in 2011 (intercept) and rate of change from 2011-2017 (slope) by REGION (4).

```{r region_int_slope}
kable(
  data %>% distinct(region, region_int, region_slope)
)
```

<br>

#### As seen in one-way ANOVA results below, there are significant differences between the region's intercept coefficient. Follow-up Tukey HSD tests indicate differences between West/Midwest, South/Northeast, and West/South. The results are visualized in the boxplot below.

```{r region_int_analysis}
summary(
  aov(state_int ~ region, data = data %>% distinct(state, region, state_int))
)

TukeyHSD(
  aov(state_int ~ region, data = data %>% distinct(state, region, state_int))
)

data %>% distinct(state, region, state_int) %>% 
  ggplot(., mapping = aes(x = region, y = state_int, color = region)) + 
    geom_boxplot() + 
    coord_flip() + 
    guides(color = FALSE) + 
    xlab("Region") + 
    ylab("State Intercept (b0)") + 
    ggtitle("State Intercept Coefficients by Region")
```

<br>

#### As seen in one-way ANOVA results below, there are no differences between the region's slope coefficient. Therefore, no follow-up analysis is needed. The results are visualized in the boxplot below.

```{r region_slope_analysis}
summary(
  aov(state_slope ~ region, data = data %>% distinct(state, region, state_slope))
)

data %>% distinct(state, region, state_slope) %>% 
  ggplot(., mapping = aes(x = region, y = state_slope, color = region)) + 
    geom_boxplot() + 
    coord_flip() + 
    guides(color = FALSE) + 
    xlab("Region") + 
    ylab("State Slope (b1)") + 
    ggtitle("State Slope Coefficients by Region")
```

<br>

#### Summary of state_int.

```{r int_summary}
summary(data %>% distinct(state, state_int) %>% select(state_int))
```

<br>

#### Summary of state_slope.

```{r slope_summary}
summary(data %>% distinct(state, state_slope) %>% select(state_slope))
```

<br>

#### A simple line plot grouped by DIVISION (sub-region) shows more variability than when grouped by REGION.

```{r division_line_plot}
data %>% select(state, division, year, percent) %>% 
  group_by(division, year) %>% 
  summarize(average = mean(percent)) %>% 
  ggplot(., mapping = aes(x = year, y = average, color = division, group = division)) + 
    geom_point() + 
    geom_line() + 
    xlab("Year") + 
    ylab("Current Cigarette Smokers (%)") + 
    ggtitle("Smoking Rates (%) by US Division") + 
    scale_x_continuous(breaks = pretty(c("2011":"2017"), n = 7))
```

<br>

#### The table below shows approximate smoking rates in 2011 (intercept) and rate of change from 2011-2017 (slope) by DIVISION (9).

```{r division_int_slope}
kable(
  data %>% distinct(division, division_int, division_slope)
)
```

<br>

#### You can see even more variability in each state's smoking trends over 2011-2017.

```{r state_line_plot}
ggplot(data, mapping = aes(x = year, y = percent, color = state, group = state)) + 
  geom_point() + 
  geom_line() + 
  xlab("Year") + 
  ylab("Current Cigarette Smokers (%)") + 
  ggtitle("Smoking Rates (%) by State") + 
  scale_x_continuous(breaks = pretty(c("2011":"2017"), n = 7)) + 
  theme(legend.position = "none")
```

<br>

#### Top 5 STATE intercepts (highest smoking rates in 2011).

```{r state_int_high}
kable(
  data %>% distinct(state, state_int) %>% 
    arrange(desc(state_int)) %>% 
    head(., 5)
)
```

<br>

#### Bottom 5 STATE intercepts (lowest smoking rates in 2011).

```{r state_int_low}
kable(
  data %>% distinct(state, state_int) %>% 
    arrange(state_int) %>% 
    head(., 5)
)
```

<br>

#### Largest (most negative) STATE slopes (greatest rate of change in reducing smokers 2011-2017).

```{r state_slope_high}
kable(
  data %>% distinct(state, state_slope) %>% 
    arrange(state_slope) %>% 
    head(., 5)
)
```

<br>

#### Smallest (least negative) STATE slopes (smallest rate of change in reducing smokers 2011-2017).

```{r state_slope_low}
kable(
  data %>% distinct(state, state_slope) %>% 
    arrange(desc(state_slope)) %>% 
    head(., 5)
)
```

<br>

#### Individual state trajectories (slope) may be influenced by where they start (intercept). Here, is a regression model showing a nonsignificant (p>0.05) negative relationship between slope and intercept for each state. The plot generated below suggests a trend such that a greater intercept results in lower [more negative] slope. This is confirmed by the negative 'state_int' estimate in the model summary.

```{r corr_int_slope}
summary(
  lm(state_slope ~ state_int, data = data %>% distinct(state, state_int, state_slope))
)

data %>% distinct(state, state_int, state_slope) %>% 
  ggplot(., mapping = aes(x = state_int, y = state_slope)) + 
    geom_point() + 
    stat_smooth(method = "lm") +
    xlab("Intercept (b0)") + 
    ylab("Slope (b1)") +
    ggtitle("Correlation between state intercept and slope coefficients")
```

<br>

### To wrap up, we will investigate factors that influence state intercept and slope coefficients.

<br>

#### TAXATION. Taxes in 2011 were significantly negatively correlated with the intercept coeffienct (top). However, change in taxation from 2011-2017 did not influence change in smoking rates (bottom).

```{r tax_intercept}
summary(
  lm(state_int ~ tax_2011, data = data %>% distinct(state, state_int, tax_2011))
)

data %>% distinct(state, state_int, tax_2011) %>% 
  ggplot(., mapping = aes(x = tax_2011, y = state_int)) + 
    geom_point() + 
    stat_smooth(method = "lm") +
    xlab("Taxation in 2011 ($ per carton)") + 
    ylab("Intercept (b0)") +
    ggtitle("Correlation between state intercept and taxation in 2011")
```

```{r tax_slope}
summary(
  lm(state_slope ~ tax_2011 + tax_2017, data = data %>% distinct(state, state_slope, tax_2011, tax_2017))
)

data %>% distinct(state, state_int, delta_tax) %>% 
  ggplot(., mapping = aes(x = delta_tax, y = state_int)) + 
    geom_point() + 
    stat_smooth(method = "lm") +
    xlab("Change in Tax from 2011-2017 ($ per carton)") + 
    ylab("Slope (b1)") +
    ggtitle("Correlation between state slope and change in taxation 2011-2017")
```

<br>

#### PREEMPTION. Number of preemptions did not have any effect on the intercept (top) or slope (bottom) state coefficients.

```{r preemption_intercept}
summary(
  aov(state_int ~ preemptions, data = data %>% distinct(state, state_int, num_of_preemptions) %>% mutate(preemptions = as.factor(num_of_preemptions)))
)

data %>% distinct(state, state_int, num_of_preemptions) %>% mutate(preemptions = as.factor(num_of_preemptions)) %>% 
  ggplot(., mapping = aes(x = preemptions, y = state_int, color = preemptions)) + 
    geom_boxplot() + 
    coord_flip() + 
    guides(color = FALSE) + 
    xlab("Number of Preemptions (#)") + 
    ylab("State Intercept (b0)") + 
    ggtitle("State Intercept Coefficients by Number of Preemptions")
```

```{r preemption_slope}
summary(
  aov(state_slope ~ preemptions, data = data %>% distinct(state, state_slope, num_of_preemptions) %>% mutate(preemptions = as.factor(num_of_preemptions)))
)

data %>% distinct(state, state_slope, num_of_preemptions) %>% mutate(preemptions = as.factor(num_of_preemptions)) %>% 
  ggplot(., mapping = aes(x = preemptions, y = state_slope, color = preemptions)) + 
    geom_boxplot() + 
    coord_flip() + 
    guides(color = FALSE) + 
    xlab("Number of Preemptions (#)") + 
    ylab("State Slope (b1)") + 
    ggtitle("State Slope Coefficients by Number of Preemptions")
```

<br>

#### INDOOR SMOKING - focus on bar restrictions. Smoking prevalence at baseline (intercept) were significantly lower in states with full restrictions on smoking in bars, although if had no effect on change in smoking prevalence over time.

```{r indoor_intercept}
summary(
  aov(state_int ~ bar_restrictions, data = data %>% 
                                            distinct(state, state_int, bar_restrictions) %>% 
                                            filter(bar_restrictions != "Designated Areas", 
                                                   bar_restrictions != "Separate Ventilated Areas"))
)

data %>% distinct(state, state_int, bar_restrictions) %>% filter(bar_restrictions != "Designated Areas", 
                                                                 bar_restrictions != "Separate Ventilated Areas") %>% 
  ggplot(., mapping = aes(x = bar_restrictions, y = state_int, color = bar_restrictions)) + 
    geom_boxplot() + 
    coord_flip() + 
    guides(color = FALSE) + 
    xlab("Bar Restriction") + 
    ylab("State Intercept (b0)") + 
    ggtitle("State Intercept Coefficients by Bar Restriction")
```

```{r indoor_slope}
summary(
  aov(state_slope ~ bar_restrictions, data = data %>% 
                                            distinct(state, state_slope, bar_restrictions) %>% 
                                            filter(bar_restrictions != "Designated Areas", 
                                                   bar_restrictions != "Separate Ventilated Areas"))
)

data %>% distinct(state, state_slope, bar_restrictions) %>% filter(bar_restrictions != "Designated Areas", 
                                                                   bar_restrictions != "Separate Ventilated Areas") %>% 
  ggplot(., mapping = aes(x = bar_restrictions, y = state_slope, color = bar_restrictions)) + 
    geom_boxplot() + 
    coord_flip() + 
    guides(color = FALSE) + 
    xlab("Bar Restriction") + 
    ylab("State Slope (b1)") + 
    ggtitle("State Slope Coefficients by Bar Restriction")
```

<br>

#### FUNDING - focus on expenditures per capita. Funding did not have any effect on the intercept (top) or slope (bottom) state coefficients. However, trends indicate that states with a higher smoker prevalence and states with a smaller rate of smoking decline spent more on smoking cessation programs.

```{r funding_intercept}
summary(
  lm(state_int ~ expenditures_per_capita, data = data %>% distinct(state, state_int, expenditures_per_capita))
)

data %>% distinct(state, state_int, expenditures_per_capita) %>% 
  ggplot(., mapping = aes(x = expenditures_per_capita, y = state_int)) + 
    geom_point() + 
    stat_smooth(method = "lm") +
    xlab("Expenditures per capita ($)") + 
    ylab("Intercept (b0)") +
    ggtitle("Correlation between state intercept and State Expenditures on Smoking Cessation")
```

```{r funding_slope}
summary(
  lm(state_slope ~ expenditures_per_capita, data = data %>% distinct(state, state_slope, expenditures_per_capita))
)

data %>% distinct(state, state_slope, expenditures_per_capita) %>% 
  ggplot(., mapping = aes(x = expenditures_per_capita, y = state_slope)) + 
    geom_point() + 
    stat_smooth(method = "lm") +
    xlab("Expenditures per capita ($)") + 
    ylab("Slope (b1)") +
    ggtitle("Correlation between state slope and State Expenditures on Smoking Cessation")
```

<br>

#### MEDICAID - focus on fee-for-service plans. No differences were observed based on medicaid services provided.

```{r medicaid_intercept}
summary(
  aov(state_int ~ fee_for_service_plans, data = data %>% distinct(state, state_int, fee_for_service_plans) %>% filter(fee_for_service_plans != "Not Applicable"))
)

data %>% distinct(state, state_int, fee_for_service_plans) %>% 
  filter(fee_for_service_plans != "Not Applicable") %>%
  ggplot(., mapping = aes(x = fee_for_service_plans, y = state_int, color = fee_for_service_plans)) + 
    geom_boxplot() + 
    coord_flip() + 
    guides(color = FALSE) + 
    xlab("Does State Provide Service?") + 
    ylab("State Intercept (b0)") + 
    ggtitle("State Intercept Coefficients by Fee-For-Service Plan")
```

```{r medicaid_slope}
summary(
  aov(state_slope ~ fee_for_service_plans, data = data %>% distinct(state, state_slope, fee_for_service_plans) %>% filter(fee_for_service_plans != "Not Applicable"))
)

data %>% distinct(state, state_slope, fee_for_service_plans) %>% 
  filter(fee_for_service_plans != "Not Applicable") %>%
  ggplot(., mapping = aes(x = fee_for_service_plans, y = state_slope, color = fee_for_service_plans)) + 
    geom_boxplot() + 
    coord_flip() + 
    guides(color = FALSE) + 
    xlab("Does State Provide Service?") + 
    ylab("State Slope (b1)") + 
    ggtitle("State Slope Coefficients by Fee-For-Service Plan")
```

<br>

#### QUITLINE. As was shown previously, states with a higher smoking prevalence (intercept) in 2011 had more quitline calls per 10,000. There was no effect of quitline calls on change in smoking prevalence over time.

```{r quitline_intercept}
summary(
  lm(state_int ~ quitline_calls_per_10000, data = data %>% distinct(state, state_int, quitline_calls_per_10000))
)

data %>% distinct(state, state_int, quitline_calls_per_10000) %>% 
  ggplot(., mapping = aes(x = quitline_calls_per_10000, y = state_int)) + 
    geom_point() + 
    stat_smooth(method = "lm") +
    xlab("Quitline calls (#/10000)") + 
    ylab("Intercept (b0)") +
    ggtitle("Correlation between state intercept and Quitline Calls")
```

```{r quitline_slope}
summary(
  lm(state_slope ~ quitline_calls_per_10000, data = data %>% distinct(state, state_slope, quitline_calls_per_10000))
)

data %>% distinct(state, state_slope, quitline_calls_per_10000) %>% 
  ggplot(., mapping = aes(x = quitline_calls_per_10000, y = state_slope)) + 
    geom_point() + 
    stat_smooth(method = "lm") +
    xlab("Quitline calls (#/10000)") + 
    ylab("Slope (b1)") +
    ggtitle("Correlation between state slope and Quitline Calls")
```

<br>

#### Information about the packages used, their versions, and the version of R that was used.

```{r session_info}
devtools::session_info()
```
